Ten Simple Rules for Teaching Programming

### 2017-07-10 paper.tex
One more pass through the paper.

\documentclass{article}
\usepackage[a4paper]{geometry}
\usepackage{natbib}
\usepackage{hyperref}

\title{Ten Simple Rules for Teaching Programming}
\author{Neil C. C. Brown\footnote{King's College London / neil.c.c.brown@kcl.ac.uk} and Greg Wilson\footnote{Rangle.io / gvwilson@third-bit.com}}

@@

One-on-one tutoring is the ideal form of teaching:
all of a teacher's attention can be focused on one student,
and they can completely customise their teaching for that person
and tailor individual feedback and corrections based on two-way dialogue with them.
In realistic settings,
however,
one teacher must teach a handful, tens, or hundreds of students at once.
How can teachers possibly hope to clear up many learners' different misconceptions
in these larger settings in a reasonable time?

@@

\item The instructor then gives learners a multiple choice question
  that probes for misconceptions rather than simple factual recall.
  A programming example is given in figure \ref{fig:mcq}
  which relates to integer comparison and loops.

@@

\begin{figure}
\begin{verbatim}
for (int i = 1; i < 10; i++) {
    if (i < 3 || i > 8) {
        System.out.println("Yes");
    }
}
\end{verbatim}
How many times will the above code print out the word Yes?\\
a) 10\\
b) 9\\
c) 4\\
d) 3
\label{fig:mcq}
\caption{An example multiple choice question probing learners' understanding of loops and integer comparisons.}
\end{figure}

@@

A good idea when teaching is
to provide many similar examples for learners to practice on.
But since learners are novices,
they may not see the similarity between examples:
finding the highest rainfall from a list of numbers and
finding the first surname alphabetically from a list of names
may seem like quite different problems to them,
even though more advanced programmers would recognise them as isomorphic.

\citet{morrison-subgoals,morrison-parsons,margulieux} have shown that students perform better
when worked examples are broken down into steps (or subgoals)
which are given names (or labels) (Figure~\ref{fig:subgoal}).
Subgoal labels provide a structure which students can apply to future tasks that they attempt themselves,
and allow them to communicate with their peers and instructors more efficiently.

\begin{figure}

\noindent
\textbf{Conventional Materials}
\begin{verbatim}
1. Click on "My Blocks" to see the blocks for components you created.
2. Click on "clap"
3. Drag out a when clap.Touched block
4. Click on “clapSound”
5. Drag out call clapSound.Play
6. Connect it after when clap.Touched
\end{verbatim}

@@

4. Click on “clapSound” and
5. Drag out call clapSound.Play
6. Connect it after when clap.Touched
\end{verbatim}
\label{fig:subgoal}
\caption{An example of subgoal labeling (taken from \cite{margulieux})}
\end{figure}

@@

One caution is that context can inadvertently exclude some people while drawing others in.
For example,
many educators use computer games as a motivating example for programming classes,
but some learners may associate them with violence and racial or gender stereotypes,
or simply find them unenjoyable.
Whatever examples are chosen,
the goal must be to move learners as quickly as possible
from ``hard and boring'' to ``easy and exciting'' \cite{repenning}.

@@

Novices do not become professionals simply by doing what professionals do at a slower pace.
We do not teach reading by taking a classic novel and simply proceeding more slowly.
We teach by using shorter books with simpler words and larger print.
So in programming,
we must take care to use small, self-contained tasks at a level suitable for novices,
with tools that suit their needs.

@@

Like anything involving human subjects,
studies of computing education must necessarily be hedged with qualifiers.
However,
we do know a great deal,
and are learning more each year.
Venues like SIGCSE (\url{http://sigcse.org/}) and ITiCSE (\url{http://iticse.acm.org/})
present a growing number of rigorous, insightful studies
with immediate practical application.
Future work may overturn or qualify some of our ten simple rules,
but they form a solid basis for any educational effort.

>>>>

\documentclass{article}
\usepackage[a4paper]{geometry}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{mdframed}

\title{Ten Simple Rules for Teaching Programming}
\author{Greg Wilson\footnote{Rangle.io / gvwilson@third-bit.com} and Neil C. C. Brown\footnote{King's College London / neil.c.c.brown@kcl.ac.uk}}

@@

One-on-one tutoring is perhaps the ideal form of teaching:
all of a teacher's attention can be focused on one student,
and they can completely customise their teaching for that person
and tailor individual feedback and corrections based on two-way dialogue with them.
In realistic settings,
however,
one teacher must usually teach several, tens, or even hundreds of students at once.
How can teachers possibly hope to clear up many learners' different misconceptions
in these larger settings in a reasonable time?

@@

\item The instructor then gives learners a multiple choice question
  that probes for misconceptions rather than simple factual recall.
  A programming example is given in figure \ref{fig:mcq}
  which relates to integer comparison and loops.  There is no point asking a trivial question
  that all students will get right.  The ideal questions are those where students are most likely to vote equally for each of the options.

@@

\begin{figure}
\begin{mdframed}
\begin{verbatim}
for (int i = 1; i < 10; i++) {
    if (i < 3 || i >= 8) {
        System.out.println("Yes");
    }
}
\end{verbatim}
How many times will the above code print out the word Yes?\\
a) 10\\
b) 5\\
c) 4\\
d) 3
\label{fig:mcq}
\caption{An example multiple choice question probing learners' understanding of loops and integer comparisons.}
\end{mdframed}
\end{figure}

@@

Instructors usually provide many similar programming examples for learners to practice on.
But since learners are novices,
they may not see the similarity between examples:
finding the highest rainfall from a list of numbers and
finding the first surname alphabetically from a list of names
may seem like quite different problems to them,
even though more advanced programmers would recognise them as isomorphic.

\citet{morrison-subgoals,morrison-parsons,margulieux} have shown that students perform better
when worked examples are broken down into steps (or subgoals)
which are given names (or labels) -- an example is given in figure~\ref{fig:subgoal}.
Subgoal labels provide a structure which students can apply to future tasks that they attempt themselves,
and allow them to communicate with their peers and instructors more efficiently.

\begin{figure}
\begin{mdframed}
\noindent
\textbf{Conventional Materials}
\begin{verbatim}
1. Click on "My Blocks" to see the blocks for components you created.
2. Click on "clap"
3. Drag out a when clap.Touched block
4. Click on “clapSound”
5. Drag out call clapSound.Play
6. Connect it after when clap.Touched
\end{verbatim}

@@

4. Click on “clapSound” and
5. Drag out call clapSound.Play
6. Connect it after when clap.Touched
\end{verbatim}
\label{fig:subgoal}
\caption{An example of subgoal labeling (taken from \cite{margulieux})}
\end{mdframed}
\end{figure}

@@

One caution is that context can inadvertently exclude some people while drawing others in.
For example,
many educators use computer games as a motivating example for programming classes,
but some learners may associate them with violence and racial or gender stereotypes,
or simply find them unenjoyable.
Whatever examples are chosen,
the goal must be to move learners as quickly as possible
from ``hard and boring'' to ``easy and exciting'' \citep{repenning}.

@@

Novices do not become professionals simply by doing what professionals do at a slower pace.
We do not teach reading by taking a classic novel and simply proceeding more slowly.
We teach by using shorter books with simpler words and larger print.
So in programming,
we must take care to use small, self-contained tasks at a level suitable for novices,
with tools that suit their needs, and without scoffing.

@@

Like anything involving human subjects,
studies of computing education must necessarily be hedged with qualifiers.
However,
we do know a great deal,
and are learning more each year.
Venues like SIGCSE (\url{http://sigcse.org/}), ITiCSE (\url{http://iticse.acm.org/}) and ICER (\url{https://icer.hosting.acm.org})
present a growing number of rigorous, insightful studies
with immediate practical application.
Future work may overturn or qualify some of our ten simple rules,
but they form a solid basis for any educational effort.

We offer one final observation: do not forget the human element.  Programmers have a reputation for pouring scorn on certain programming languages (e.g. PHP), or for gatekeeping (e.g. stating that you can't learn programming if you didn't start young).  If you are teaching someone to program, the last thing you want to do is make them feel like they can't succeed or that any existing skill (perhaps in PHP) they have is worthless.  Make your learners feel that they can be a programmer, and they just might become one.
